Title Goes Here

### 2021-02-26 analysis/paper/paper.Rmd
Más PSA SAM y texto

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    fig_height: 7.6
    fig_width: 12.8
  powerpoint_presentation:
    fig_height: 7.6
    fig_width: 12.8
    reference_doc: ../templates/template.pptx
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
header-includes:
  - \usepackage{subfig}
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 40,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

@@

guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
  guide_colorsteps(show.limits = TRUE, even.steps = FALSE,
                   title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

geom_contour_pval <- function(mapping, p.value = 0.01, size = 0.1) {
  list(
    stat_contour_fill(mapping, breaks = c(0, p.value), fill = NA, geom = pattern_dots),
    geom_contour2(mapping, breaks = p.value, size = size)
  )
}

@@

```{r o3}
o3 <- O3() %>% 
  readRDS() %>% 
  .[lat <= -20] %>% 
  .[, .(toc = mean(total_ozone_column, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))] 
o3[, time := year(time)]
```

@@

@@

@@

@@

ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30")
```

Both observations motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time. Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and argument of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the navie EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3. Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean. The real part of the first EOF is, therefore, capturing the mean zonally anomalous field and the variability that projects onto it.

@@

  facet_grid(PC ~ part, labeller = labeller(PC = lab_PC), scales = "free_y") +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  tag_facets("cr")
```

@@

## Regression

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt")
```

Figure \@ref(fig:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.

@@

```{r zonal-waves, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
regr %>%
  .[lat <= -20] %>%
  .[lev %in% c(200, 50)] %>%
  # .[PC == "PC1"] %>%
  .[, FitWave(estimate, 0:3), by = .(lev, PC, term, lat)] %>%
  ggplot(aes(lat, amplitude)) +
  geom_line(aes(color = factor(k))) +
  ggh4x::facet_nested(lev + PC ~ term, scales = "free") +
  scale_color_brewer("Wavenumber", palette = "Dark2") +
  scale_y_continuous("Amplitude") +
  scale_x_latitude() +
  coord_flip()
```

Interpretaciones:

-   Parte real de EOF1: variabilidady que se proyecta en la dirección del campo medio

-   Parte imaginaria de EOF1: variabilidad asociada a la parte simétrica del SAM

-   EOF2: PSA1 y PSA2.\

-   Parte imaginaria: proyecta sobre la parte asimétrica del SAM (PSA2)

-   Parte real: relativamente independiente del SAM

El primer EOF es esencialmente una medida de la variabilidad proyectada en la dirección el campo medio. La parte importante, de todas formas, es la Real.

@@

(ref:eof2-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second EOF and temperature.

```{r t}
t <- ERA5() %>% 
  ReadNetCDF(vars = "t", 
             subset = list(latitude = c(-90, -20))) %>% 
  normalise_coords()
```

```{r t-vertical, fig.cap = "Regression of EOF1 with mean air temperature between 80ºS and 25ºS."}
# t_PC1_vertical <- t %>% 
#   .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
#   sep_ReIm(format = "wider") %>% 
#   .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon, lat)] %>% 
#   rm_intercept()
# t_PC1_vertical %>% 
#   .[lev %in% c(1, 5, 10, 50)] %>% 
#   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
#     ggplot(aes(lon, lat)) +
#   geom_contour_fill(aes(z = estimate)) +
#   stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)),
#               size = .5, alpha = 0.5) +
#   geom_hline(yintercept = c(-80, -35)) +
#   geom_qmap(~.x[lat <= -20]) +
#   scale_fill_divergent() + 
#   coord_polar() +
#   facet_grid(lev ~ term)

t %>% 
  .[lat %between% c(-80, -35)] %>% 
  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>%
  ggplot(aes(lon, lev)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), 
                      breaks = AnchorBreaks(0, exclude = 0)) +

  geom_contour_pval(aes(z = p.value)) +
  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom()) + 
  scale_y_level() +
  scale_x_longitude() +
  facet_grid(~term) +
  axis_labs_smol
```

@@

```{r o3-regr, fig.cap = "Regression of standarised EOF1 and Total Ozone column."}
o3_regr %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +
  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   guide = guide_colorsteps_bottom()) + 
  scale_x_longitude() +
  scale_y_latitude() +
  coord_polar() +
  facet_grid(PC~term) +
  axis_labs_smol
```

The ozone hole, as measured by Total Ozone Column (TOC) is climatologically not centered over the South Pole, but shifted towards the Weddell Sea, which translates to a large amplitude zonal wave 1 with minimum over the Weddell Sea and maximum over the Davis Sea [@wirth1993]. The regression

@@

temp %>% 
  copy() %>% 
  na.omit() %>% 
  sep_ReIm() %>% 
  .[, hgt := scale(hgt), by  = .(part)] %>% 
  ggplot(aes(amplitude, hgt)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(PC ~ part)
```

```{r eof2-regr-t, fig.cap = "(ref:eof2-regr-t-cap)"}
plot_regr(regr, "PC2", c(200, 50), "t")
```

## Relationship with other indices

### PSA1 and PSA2

@@

pc2[psa_time, on = "time"] %>% 
  .[PC != "SAM"] %>% 
  copy() %>% 
  # .[, hgt := rotate(hgt, psa_angle)] %>%
  sep_ReIm() %>% 
  ggplot(aes(value, hgt)) +
  geom_point() +
  scale_y_continuous("EOF2") +
  scale_x_continuous("PSA") +
  facet_grid(part ~ PC)
```

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021. Points mark estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].

@@

sam_r2 %>%
  ggplot(aes(lev, estimate^2)) +
  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
  geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²", limits = c(0, 1)) +
  scale_size_manual(guide = "none",
                    values = c(full = 1.2,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "A-SAM",
                                                         sym = "S-SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip() +
  tag_facets("rc")
```

@@

This suggests that the separation between the asymmetric and symmetric part of the SAM is not only statistically possible, but physically plausible. The asymmetric part of the SAM might be statistical contamination from one phase of the PSA mode.

>>>>

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    fig_height: 7.6
    fig_width: 12.8
  powerpoint_presentation:
    fig_height: 7.6
    fig_width: 12.8
    reference_doc: ../templates/template.pptx
bibliography: [references.bib, data.bib]
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
header-includes:
  - \usepackage{subfig}
# abstract: |
#   Text of abstract
# keywords: |
#   keyword 1; keyword 2; keyword 3
# highlights: |
#   These are the highlights. 
---

@@

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE,
  comment = "#>",
  cache.extra = 41,
  cache.path = "../cache/paper/",
  fig.path = "../figures/",
  dpi = 300,
  dev = c("pdf", "png")
)

@@

panel_background <- theme(panel.background = element_rect(fill = "#fbfbfb", color = NA), 
                          panel.ontop = FALSE,
                          tagger.panel.tag.background = ggplot2::element_rect(color = NA, 
                                                                              fill = "#fbfbfb"))

@@

guide_colorsteps_bottom <- function(width = 25, height = 0.5, even.steps = FALSE, ...) {
  guide_colorsteps(show.limits = TRUE, even.steps = even.steps,
                   title.position = "top", title.hjust = 0.5,
                   barheight = height,
                   barwidth = width, ...)
}

@@

geom_contour_pval <- function(mapping, p.value = 0.01, size = 0.1) {
  mapping2 <- mapping
  mapping2$fill <- aes(fill = NA)$fill
  list(
    stat_contour_filled(mapping2, breaks = c(0, p.value), fill = NA, geom = pattern_dots),
    geom_contour2(mapping, breaks = p.value, size = size)
  )
}

@@

```{r o3}
o3 <- O3() %>% 
  readRDS() %>% 
  .[lat <= -20] %>% 
  .[, .(toc = mean(total_ozone_column, na.rm = TRUE)), by = .(lon, lat, time = seasonally(time))] %>% 
  unique(by = c("lon", "lat", "time"))
o3[, time := year(time)]
```

@@

@era5 and @era5be

@@

## Motivation to use complex EOF

@@

Both observations motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.

## Description of EOFs

@@

ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the navie EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3. Note that in Figure \@ref(fig:ceofs-2), both parts of the EOF1 have non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean. The variability associated with the first EOF is variability that projects onto the mean zonally anomalous field.

@@

  facet_grid(PC ~ part, labeller = labeller(PC = lab_PC), scales = "free_y") +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  tag_facets("cr") +
  panel_background
```

@@

## Regressions

The spatial patterns shown in Figure \@ref(fig:ceofs-1) are "pure" EOF patterns. But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series. They are also idealised in that they are forced by construction, to be Hilber transforms of each other. To get a more realistic view of the real-world variability described by these EOFS, we computed regression patterns of each EOF with the complete geopotential fields.

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt") +
  # geom_contour2(data = ~copy(.x)[, estimate := Anomaly(estimate), by = .(term, lev, lat)],
  #              aes(z = estimate, linetype = factor(-sign(..level..))), 
  #              global.breaks = FALSE, breaks = AnchorBreaks(0, bins = 2, exclude = 0)) +
  # scale_linetype(guide = "none") +
  NULL
```

Figure \@ref(fig:eof1-regr-gh) shows regression patterns of EOF1 and geopotential height. At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica. However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole. 
Similarly affected are the regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b). It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure @\ref(fig:eof1-regr-gh)b.1). The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.

With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure \@ref(fig:ceofs-1), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa, suggesting that using the complex EOF method is "forcing" a wave structure where there is none.

Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere they are associated with slightly off-centre monopoles.

@@

Figure \@ref(fig: eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2. Unlike in the case of EOF1, in this case the regression patterns represent relatively well the *idealised* patterns from Figure \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere.

EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001]. Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig: eof2-regr-gh) with Figure 1 a and b from @mo2001, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1. This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.

```{r zonal-waves, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
# regr %>%
#   .[lat <= -20] %>%
#   .[lev %in% c(200, 50)] %>%
#   # .[PC == "PC1"] %>%
#   .[, FitWave(estimate, 0:3), by = .(lev, PC, term, lat)] %>%
#   # .[, pseudor2 := amplitudepseudor2^2/sum(amplitude^2), by = .(lev, PC, term, lat)] %>% 
#   ggplot(aes(lat, amplitude)) +
#   geom_line(aes(color = factor(k))) +
#   ggh4x::facet_nested(PC + lev ~ term, scales = "free", labeller = labeller(lev = lab_lev)) +
#   scale_color_brewer("Wavenumber", palette = "Dark2") +
#   scale_y_continuous("Amplitude") +
#   scale_x_latitude() +
#   coord_flip() +
#   panel_background
```

## Relationship with other variables

### Ozone

@@

```{r t}
t <- ERA5() %>% 
  ReadNetCDF(vars = "t", 
             subset = list(latitude = c(-90, -20))) %>% 
  normalise_coords() %>% 
    .[season(time) == "SON"] %>%
  .[, .(t = mean(t)), by = .(lev, lon, lat, time = seasonally(time))]
```

```{r t-vertical, fig.cap = "Regression of EOF1 with mean air temperature between 80ºS and 25ºS."}
t %>% 
  .[lat %between% c(-80, -35)] %>% 
  .[, .(t = weighted.mean(t, cos(lat*pi/180))), by = .(lev, time, lon)] %>%  
  .[, t := Anomaly(t), by = .(time, lev)] %>% 
  .[cut(ceof$eof[[1]], 1)$left, on = "time"] %>% 
  sep_ReIm(format = "wider") %>% 
  .[, ":="(Real = as.numeric(scale(Real)), Imaginary = as.numeric(scale(Imaginary))), by = .(PC)] %>% 
  .[, FitLm(t, Real, Imaginary, se = TRUE), by = .(lev, lon)] %>% 
  rm_intercept() %>% 
  .[, term := factor_ReIm(term)] %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>%
  ggplot(aes(lon, lev)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), 
                      breaks = AnchorBreaks(0, 0.25, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +
  scale_fill_divergent_discretised("Air temperature", guide = guide_colorsteps_bottom(25)) + 
  scale_y_level() +
  scale_x_longitude() +
  facet_grid(~term) +
  axis_labs_smol
```

Figure \@ref(fig:eof1-regr-t) shows regression patterns of EOF1 with air temperature. In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa. To understand the vertical distribution of these anomalies, Figure @\ref(fig:t-vertical)) shows the regression pattern of EOF1 and zonal anomalies of air temperature averaged between 80ºS and 35ºS. Both pars of EOF1 are associated with a temperature wave 1 pattern which changes phase around 10 hPa. This switch in the sign of the correlation in temperature at 10 hPa suggest a relationship with Ozone, since its indicative of the effects of planetary waves in the advection of Ozone [@hartmann1979;@rood1985;@smith1995].

@@

```{r o3-regr, fig.cap = "Regression of standarised EOF1 and Total Ozone column."}
o3_regr %>% 
  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(PC, term)] %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate, fill = ..level..),
                    breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = estimate), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_pval(aes(z = p.value)) +
  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   guide = guide_colorsteps_bottom()) + 
  scale_x_longitude() +
  scale_y_latitude() +
  coord_polar() +
  facet_grid(PC~term) +
  axis_labs_smol +
  tag_facets("panel")
```

To confirm this possibility, Figure \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC). It shows zonal wave 1 pattens in TOC associated with both phases of EOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993]. Thus, the Real EOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º. This suggest that the EOF1 can be a complete description of the TOC zonal wave 1.

@@

# temp %>% 
#   copy() %>% 
#   na.omit() %>% 
#   sep_ReIm() %>% 
#   .[, hgt := scale(hgt), by  = .(part)] %>% 
#   ggplot(aes(amplitude, hgt)) +
#   geom_point() +
#   geom_smooth(method = "lm", color = "black") +
#   facet_grid(PC ~ part) +
#   panel_background
# temp %>% 
#   .[, correlate(amplitude, Mod(hgt))]
# 
# temp %>% 
#   .[, correlate(phase, Arg(hgt))]

correlation_statistics <- function(x, n, signif = 2) {
  z <- atanh(x)
  se <- 1/sqrt(n - 3)

  low <- qnorm(.025, mean = z, sd = se)
  low <- tanh(low)
  hig <- qnorm(.975, mean = z, sd = se)
  hig <- tanh(hig)

  p.value <- pnorm(abs(x)/se, lower.tail = FALSE)

  text <- paste0(signif(x, signif), " (CI: ",
                 signif(low, signif), " -- ", signif(hig, signif), ")")

  list(estimate = x, 
       p.value = p.value, 
       low = low,
       hig = hig,
       text = text
  )
}

wave_correlation <- temp %>% 
  na.omit() %>% 
  .[, .(cor = mean(cos(phase - Arg(hgt))),
        n = .N)] %>% 
  .[, correlation_statistics(cor, n)]

temp %>% 
  ggplot(aes(Mod(hgt), amplitude)) +
  scale_x_continuous("EOF1 Amplitude") +
  scale_y_continuous("O3 Wave-1 Amplitude") +

temp %>% 
  copy() %>%
  na.omit() %>% 
  qwrap(phase = c(0, 2*pi) ~ c(-pi, pi)) %>% 
  ggplot(aes(Arg(hgt), phase)) +
  scale_x_continuous("EOF1 Phase",
                     breaks = seq(-pi, pi, by = 15*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  scale_y_continuous("O3 Wave-1 Phase",
                     breaks = seq(-pi, pi, by = 15*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º"))  +
  plot_layout(ncol  = 1) &
  geom_point()  &
  geom_smooth(method = "lm") 
```

Indeed, Figure \@ref(fig:o3-regr) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1. The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`. As expected from the location of anomalies in Figure \@ref(fig: o3-regr), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).

### PSA

We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components .

@@

pc2[psa_time, on = "time"] %>% 
  .[PC != "SAM"] %>% 
  copy() %>% 
  # .[, hgt := rotate(hgt, psa_angle)] %>%
  sep_ReIm() %>% 
  ggplot(aes(value, hgt)) +
  geom_point() +
  scale_y_continuous("EOF2") +
  scale_x_continuous("PSA") +
  facet_grid(part ~ PC) +
  panel_background +
  tag_facets("rc")
```

Figure @\ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2. As anticipated by Figure @\ref(fig: eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure @\ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure @\ref(fig:psa-eof2)a.2). Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure @\ref(fig:psa-eof2) panels a.1 and b.2). So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2. It is also the same that maximises this clean association between EFO2 parts and PSA modes. The reason that the "naive" principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.

```{r phase-histogram, fig.cap = "Histogram of phase distributions of EOF2."}
cut(ceof$eof[[1]], 2)$left %>% 
  .[, arg := Arg(hgt)] %>% 
  .[, w := Mod(hgt)/sum(Mod(hgt))] %>% 
  qwrap(arg = c(-pi, pi) ~  c(-pi - 1/4*pi, pi - 1/4*pi)) %>% 
  ggplot(aes(arg)) +
  geom_histogram(binwidth = 90*pi/180, center = 0, color = "black", fill = "gray70") +
  annotate("text",
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
           color = "black", size = 6,
           y = 0.5, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +
  geom_rug() +
  scale_y_continuous("Number of years") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi - 1/4*pi, pi - 1/4*pi, by = 45*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) 
```

By aligning he imaginary phase with the direction with maximum relationship with ENSO, we also aligned it with the phase of maximum occurrence. This is analogous to the method in @irving2016, who used reprojection and Fourier filtering to detect "PSA-like" variability and defined as the proper PSA using the peaks of the phase distribution (Compare our Figure \@ref:(phase-histogram) with their Figure 6).

The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillation.

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021. Points mark estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].

@@

sam_r2 %>%
  ggplot(aes(lev, estimate^2)) +
  geom_vline(xintercept = c(50, 200),  size = 0.2, color = "gray50") +
  geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²", limits = c(0, 1)) +
  scale_size_manual(guide = "none",
                    values = c(full = 1.2,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "A-SAM",
                                                         sym = "S-SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip() +
  tag_facets("rc") +
  panel_background
```

@@

```{r fake-eof, fig.cap = "Principal compomnents of a sinthetic data set of geopotential height as the sum of the fields predicted by the EOF2 plus perfectly zonally symmetric SAM-like annomalies with random amplitude."}
pattern <- eof_filtered$right %>% 
  .[era5[lev == 200], on = "time"] %>% 
  .[, FitLm(hgt, eof), by = .(lon, lat)] %>% 
  .[, estimate := mean(estimate), by = .(lat, term)] %>%
  dcast(lon + lat ~ term, value.var = "estimate")

set.seed(42)
fake_sam <- eof_filtered$right %>% 
  copy() %>% 
  .[, eof := sample(eof, replace = TRUE)]  %>%
  setnames("eof", "value") %>% 
  .[, pattern[, .(sam = value*eof), by = .(lon, lat)], by = time]

# fake_sam %>% 
#   .[time %in% unique(time)[1:4]] %>% 
#   ggplot(aes(lon, lat)) +
#   geom_contour_fill(aes(z = sam)) +
#   facet_wrap(~factor(time)) +
#   coord_polar()

fake_eof <- era5 %>% 
  .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
  .[lev == 200] %>% 
  .[lat <= -20] %>% 
  .[season(time) == "SON"] %>% 
  .[, hgt := hgt - .lm.fit(cbind(1, Imaginary, Real), hgt)$residuals, .(lon, lat, lev, season(time))] %>% 
  .[, hgt := Anomaly(hgt), by = .(lat)] %>% 
  fake_sam[., on = .NATURAL] %>% 
  .[, hgt := hgt + sam] %>% 
  .[, eof := Anomaly(hgt)*sqrt(cos(lat*pi/180)), by = .(lon, lat, season(time))] %>% 
  EOF(eof ~ lon + lat | time, n = 1:3, data = .)

# era5 %>% 
#   .[sep_ReIm(pc2, format = "wide")[, hgt := NULL], on = .NATURAL] %>% 
#   .[lev == 200] %>% 
#   .[lat <= -20] %>% 
#   .[season(time) == "SON"] %>% 
#   .[, FitLm(hgt, Real, Imaginary), .(lon, lat)] %>% 
#   rm_intercept() %>% 
#   .[, .(lon, lat, eof = estimate, PC = term)] %>% 
#   rbind(pattern[, .(lon, lat, eof, PC = "SAM")]) %>% 
#   .[, PC := factor(PC, levels = c("SAM", "Imaginary", "Real"), ordered = TRUE)] %>% 
#   periodic(lon = c(0, 360)) %>% 
#   ggplot(aes(lon, lat)) +
#   geom_contour_fill(aes(z = eof), global.breaks = FALSE, breaks = AnchorBreaks(0, exclude = 0)) +
#   geom_contour_tanaka(aes(z = eof), global.breaks = FALSE,  breaks = AnchorBreaks(0, exclude = 0)) +
#   scale_fill_divergent(guide = "none") +
#   geom_qmap(~.x[lat <= -20]) +  
#   scale_x_longitude() +
#   scale_y_latitude() +
#   coord_polar() +
#   facet_wrap(~PC) +
#   axis_labs_smol +
